\documentclass{frama-c-book}
\usepackage{graphicx}
\usepackage{calc}

\input{fc-macros}

\newcommand{\FCVERSION}{\input{VERSION}(\input{VERSION_CODENAME}\unskip)}

\title{Volatile Plug-in Manual}
\author{Patrick Baudin}
\fcversion{\FCVERSION}

\cclicence{by}
\copyrightstarts{2011}

\makeindex

\begin{document}
\sloppy
\emergencystretch 3em

\maketitle
\tableofcontents

\include{macros}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Overview}

This is the user manual of the Volatile plug-in of \FramaC\footnote{\url{http://frama-c.com}}.
The content of this document corresponds to the version \FCVERSION, released on \today, of
\FramaC.

This plug-in adds to \FramaC support for \C qualifier {\tt volatile}.
It builds a new project where volatile accesses (loads from and stores to volatile data) are simulated by function calls, according to the \acsl\cite{acsl-implem} clauses {\tt volatile}.
Therefore, it is useful to perform analyses on this new project without taking care of the {\tt volatile} qualifiers since the introduced function calls model the effects of these volatile accesses.

The detection of volatile accesses is purely syntactic.
The plug-in looks at the l-values of the source code.
\acsl\cite{acsl-implem} clauses {\tt volatile} should inform the plug-in about the functions to introduce for modeling the accesses to the l-values having the qualifier {\tt volatile} or being composed of a subpart (i.e. a field) having that {\tt volatile} qualifier.
A warning message is emitted when no function is found for such l-value.

The plug-in warns also about casts from pointers to volatile data since accesses to these data may be undetected by the syntactic search.

An instrumentation of calls through function pointers is also available.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Volatile}

You may refer to the user manual of \FramaC \cite{userman} for non-specific
options of \FramaC.

After its installation, the plug-in is listed by \verb|frama-c -plugins|
as \verb|Volatile|.

The plug-in options are:
\begin{itemize}{}
\item \verb|-volatile-help| lists all options specific to the plug-in.
\item \verb|-volatile| invokes the plug-in.
\item \verb|-volatile-fct| {\sl\/<f, ...>} specifies the list of functions to process (defaults to all functions).
\item \verb|-volatile-binding| {\sl\/<f, ...>} allows binding of the volatile accesses to the specified functions.
\item \verb|-volatile-binding-auto| allows automatic binding to functions from the type name of the volatile accesses: functions \verb|Rd_|{\sl\/<type-name>} and \verb|Wr_|{\sl\/<type-name>} for read and write accesses (the use of the option \verb|-volatile-msg-key="binding"| could help to guess the necessary {\sl <type-name>}).
\item \verb|-volatile-binding-prefix| {\sl <str>}: adds {\sl <str>} as prefix to function name for automatic binding (defaults to \verb|c2fc2_|).
\item \verb|-volatile-binding-call-pointer| replaces calls through function pointers by direct calls to functions:
  \\ {\sl <prefix>\verb|Call_|<result-type-name>\verb|_|<param-type-names>}
\item \verb|-volatile-call-pointer| {\sl <f, ...>}:  stub call to pointer functions to the provided functions (indexed by type).
\end{itemize}

The use of the plug-in could be the following:

\begin{verbatim}
> frama-c -volatile <volatile options> <C files> \
    -then-on Volatile -set-project-as-default \
    -then <analysis options>
\end{verbatim}


Here is a small example \C program contained in file \verb|demo.c|:
\cinput{demo.c}

Here is the result of the \plugin{} about the \verb|job| function:
\begin{lstlisting}[language=frama-c-commands, style=frama-c-style]
> frama-c -volatile demo.c -then-last -print
\end{lstlisting}

\begin{lstlisting}[language=c, style=frama-c-style]
int job(int x){
  int __volatile_tmp;
  wr_v(& v,x);
  __volatile_tmp = rd_v(& v);
  return __volatile_tmp;
}

\end{lstlisting}


When the plug-in is used in combination with the ACSL importer plug-in, the use could be the following:

\begin{lstlisting}[language=frama-c-commands, style=frama-c-style]
frama-c <importing options> <C files> \
  -then -volatile <volatile options> \
  -then-on Volatile <analysis options>
\end{lstlisting}

\chapter{Binding for volatile accesses}

The volatile accesses can be bound to functions from (sorted by decreasing priority order):
\begin{enumerate}
\item global \verb|volatile| clauses written into {\sl ACSL} (as into the previous example)
\item the option \verb|-volatile-binding|
\item the option \verb|-volatile-binding-auto|
\end{enumerate}
When no binding is found for an lvalue, a dedicated warning message is emitted.
However, these warnings can be silent using \verb|-volatile-no-warning-on-volatile-lvalues|.

The prototype of the functions bound using the options must be compatible to the ones specified into global \verb|volatile| clauses of {\sl ACSL}:
\begin{itemize}
\item  {\sl <T> <function-name>\verb|(volatile| <T> \verb|*)|} for read accesses to lvalues of type {\sl <T>}
\item  {\sl <T> <function-name>\verb|(volatile| <T> \verb|*, <T>)|} for write accesses to lvalues of type {\sl <T>}
\end{itemize}

For the binding feature related to the \verb|-volatile-binding|, using the
\verb|-volatile-basetype| breaks that rule since it enables a search among the
functions specified by \verb|-volatile-binding| from the base type of the volatile
access to instrument, and not the real C type.
Mainly, the type attributes and qualifiers are not considered during the search.
Of course, casts are introduced when necessary into the code instrumentation to get a compilable C code.

When using automatic binding, the option \verb|-volatile-msg-key="binding"| could help to guess the necessary {\sl <function-name>} in giving the inferred {\sl <type-name>}.

\chapter{Function Pointers}

The Volatile plug-in is also capable of instrumenting calls through function pointers.
For instance, the following statement:
\begin{verbatim}
   r = (*p)(a,b,c);
\end{verbatim}
can be instrumented by a static call to some well-chosen function \verb+F+:
\begin{verbatim}
   r = F(p,a,b,c);
\end{verbatim}

There are two mechanisms for choosing the appropriate function \verb+F+ associated to a function pointer:
\begin{itemize}
\item Either an option \verb+-volatile-call-pointer F+ is used and the prototype of \verb+F+ is compatible with
  the type of \verb+(*p)+;
\item Or the option \verb+-volatile-binding-call-pointer+ is set and \verb+F+ has a canonical name, like for volatile read/write functions.
\end{itemize}

The canonical name is \verb+<prefix>Call_<result-type>_<param-type>_..._<param-type>+, where:
\begin{itemize}
\item \verb+<prefix>+ is set by option \verb+-volatile-binding-prefix <prefix>+
\item \verb+<result-type>+ is the result type name of the call through function pointer
\item \verb+<param-type>+ is the type name of the parameter of the call (using the same order from the left to right)
\end{itemize}

\appendix
\include{user-changes}

\bibliographystyle{plain}
\bibliography{bib}
\addcontentsline{toc}{chapter}{\bibname}

\cleardoublepage
\end{document}
