%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             %
%  This file is part of the Frama-C plug-in 'ACSL importer' (acsl-importer).  %
%                                                                             %
%  Copyright (C)                                                              %
%    CEA (Commissariat à l'énergie atomique et aux énergies                   %
%         alternatives)                                                       %
%                                                                             %
%  All rights reserved.                                                       %
%  Contact CEA LIST for licensing.                                            %
%                                                                             %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{frama-c-book}
\usepackage{graphicx}
\usepackage{calc}

\input{fc-macros}

\newcommand{\FCVERSION}{\input{VERSION}(\input{VERSION_CODENAME}\unskip)}

\title{ACSL-importer Plug-in Manual}
\author{Patrick Baudin}
\fcversion{\FCVERSION}

\cclicence{by}
\copyrightstarts{2011}

\makeindex

\begin{document}
\sloppy
\emergencystretch 3em

\maketitle
\tableofcontents

\include{macros}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Overview}

This is the user manual of the \acsl\cite{acsl, acsl-implem} importer plug-in of
\FramaC\footnote{\url{http://frama-c.com}}. The content of this document
corresponds to the version \FCVERSION, released on \today, of \FramaC.

This plug-in adds to \FramaC support for out of \C source files
\acsl specification. Therefore it is useful only to people that
intend to write \acsl specifications and do not want to write them
inside comments of the \C source files as documented in \cite{acsl}.

In addition, the plug-in allows to enter \acsl annotations directly via the
Graphical User Interface.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Usage}
You may refer to the user manual of \FramaC \cite{userman} for non
specific options of \FramaC.

After its installation, the plug-in is listed by \verb|frama-c -plugins|
as \verb|ACSL importer|.

All options of the plug-in are prefixed by \verb|-acsl-import|.

The main options are:
\begin{itemize}{}
\item \verb|-acsl-import-help| lists all options specific to the plug-in
\item \verb|-acsl-import <filename list>| analyzes the content of
  files listed in its arguments and inserts the correctly typed
  specifications in the abstract syntax tree of \FramaC for further processing.
  The file name list is a comma separated list of file names.
  It can be repeated many times and its effect is cumulative.
\item \verb|-acsl-import-include-dirs <dirname list>| specifies the list of
  directories where the relative file names mentioned in \verb|include|
  directives are searched.
\item \verb|-acsl-import-keep-unused-symbols| configures \FramaC kernel such that
  unused C symbols aren't removed before the import.
\item \verb|-acsl-import-no-run| doesn't run the plug-in, just configure its parameters.
\item \verb|-acsl-import-parse-only| parses the ACSL files without typing nor imports.
\item \verb|-acsl-import-typing-only| parses and types the ACSL files without imports.
\end{itemize}

For example if you want to check your specification written in the file
named \verb|specif.acsl| for the program in \verb|example.c| using the
value analysis plug-in\cite{value}, you may use the following command:
\begin{shell}
# frama-c -acsl-import specif.acsl example.c -then -eva
\end{shell}
The option \verb|-then| ensures that the specification is imported
before running the value analysis.

\section*{Grammar for external \acsl files}
\markright{}
\addcontentsline{toc}{section}{Grammar for external \acsl files}

An external \acsl file is a text file containing any <specification> conforming to the grammar described in table \ref{grammar}.

Non terminals starting with <acsl\_> correspond to non terminals without the <acsl\_> in the original \acsl grammar\cite{acsl}.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|lrl|}
\hline
<specification> & := & (("{\tt\/module}" <module\_name> "\verb+:+")* <module\_specification>)* ;\\
           &    &\\
<module\_specification> & := &  <global\_specification>* <fun\_specification>* ;\\
           &    &\\
<global\_specification> & := & <include\_directive> | <macro\_directive>\\
           &  | & <acsl\_logical\_axiomatic> | <acsl\_logical\_type>\\
           &  | & <acsl\_logical\_predicate> | <acsl\_logical\_function>\\
           &  | & <acsl\_volatile\_clause> ;\\
           &    &\\
<include\_directive> & := & "\verb+include+" "\verb+"+"<file\_name>"\verb+"+" "\verb+;+" ;\\
           &    &\\
<macro\_directive> & := & ("\verb+global+" ?) \verb+let+" <identifier> "\verb+=+" <acsl\_term> "\verb+;+" ; \\
           &    &\\
<fun\_specification> & := & "{\tt\/function}" <fun\_name> "\verb+:+"\\
           &    & (<fun\_spec> | <stmt\_spec> | <global\_specification>)* ;\\
           &    &\\
<fun\_spec> & := & "{\tt\/contract}" "\verb+:+" <acsl\_fun\_spec> ;\\
           &    &\\
<stmt\_spec> & := & <stmt\_ref> "\verb+:+" (<acsl\_loop\_annotation>\\
           &  | &  <acsl\_code\_annotation>\\
           &  | &  "{\tt\/contract}" "\verb+:+" <acsl\_stmt\_spec>) ;\\
           &    &\\
<stmt\_ref> & := & "{\tt\/at}" (<loop\_id>\\
           &  | & <stmt\_id>  \\
           &  | & <call\_id>) ;\\
           &    &\\
<loop\_id>  & := & "{\tt\/loop}" <number> \\
           &   |& "{\tt\/loop}" "{\tt\/body}" <number>;\\
           &   |& "{\tt\/loop}" "{\tt\/stmt}" <number>;\\
           &    &\\
<stmt\_id>  & := & <number> \\
           &   |& "{\tt\/return}" \\
           &   |& <label\_name> ;\\
           &    &\\
<call\_id>  & := &"{\tt\/asm}" <number> \\
           &   |& "{\tt\/indirect\_call}" <number>* \\
           &   |& "{\tt\/call}" <fun\_name> <number>*  \\
           &   |& "{\tt\/call}" <number> ;\\

\hline
\end{tabular}
\end{center}
\caption{External \acsl grammar}
\label{grammar}
\end{table}

The importation semantics is given by the following rules:
\begin{itemize}
  \item \verb|module f : <S>| means that the specification in \verb|S|
    has to be imported into the \C file named \verb|f.c|
  \item \verb|include "f.acsl"| means that the file named
    \verb|f.acsl| has to be included verbatim into the current file,
    as if its content was inlined.
    The included file has to contain a valid \verb|<specification>|.
    If \verb|f.acsl| is a relative filename the plug-in will look for
    first inside the directory containing file containing the
    \verb|include| directive and then in first directory of the list
    of directories given to the option \verb|-acsl-import-include-dirs| that
    contains \verb|f.acsl|.
  \item \verb|let m = <T>;| defines a macro named \verb|m|. All
    occurrences of the macro \verb|m| in current scope
    will be expanded verbatim by the \acsl term \verb|T|. The
    \verb|let| is untyped but each expansion of \verb|m| into \verb|T|
    will be typed in its own context.
    The considered scope is the whole file for macro defined before the
    first \verb|module| markup, the current module before the first
    \verb|function| markup and the current function otherwise.
    Macros can be located into an included file to be shared between
    specifications of different imported files.
  \item \verb|global let m = <T>;| defines a macro named \verb|m|
    within the file scope.
  \item \verb|function f : contract : <S>| means that the
    specification in \verb|S| has to be imported as a specification
    for the \C function named \verb|f|
  \item \verb|function f : <stmt_ref> : <S>| means that the
    specification in \verb|S| has to be imported as a specification
    for the statement denoted by \verb|stmt_ref| inside the body of
    the \C function named \verb|f|
  \item \verb|at loop <i>| denotes the loop of number \verb|i|,
    inside the current function, counting from the top of the definition
    of the function and starting from $1$.
    The ACSL loop annotations refer to the loop statement itself,  whereas
    ACSL code annotations and statement contracts refer to the loop body.
  \item \verb|at loop stmt <i>| as previous, but denotes always the loop stmt
    (notice that this statement never includes the initializer part of
    \verb|for| loops),
  \item \verb|at loop body <i>| as previous, but denotes always the loop stmt
    (notice that this statement never includes the increment part of \verb|for|
    loops),
  \item \verb|at asm <i>| denotes the asm call of number \verb|i|,
     inside the current function, counting from the top of the definition
    of the function and starting from $1$.
  \item \verb|at indirect_call <i>| denotes the indirect call (via function pointer) of number \verb|i|,
    inside the current function, counting from the top of the definition
    of the function and starting from $1$.
    To refer to all indirect calls from the current function, do not give a value \verb|i|.
  \item \verb|at call <fun_name> <i>| denotes the direct call to the function \verb|fun_name| of number \verb|i|,
    inside the current function, counting from the top of the definition
    of the function and starting from $1$.
    To refer to all direct calls to  \verb|fun_name| from the current function, do not give a value \verb|i|.
  \item \verb|at call <i>| denotes the direct call of number \verb|i|,
    inside the current function, counting from the top of the definition
    of the function and starting from $1$.
  \item \verb|at <L>| denotes the statement designated by the \C label
    \verb|L|, inside the current function. In addition to the \C
    labels the special keyword \verb|return| is interpreted as
    pointing to the \verb|return| statement of the \C function.
  \item \verb|at <i>| denotes the statement of number \verb|i|, inside
    the current function, counting from the top of the definition of the
    function (in its internal form) and starting from $1$. Note that
    this number may be counter-intuitive to compute and may
    depend on the version of the \FramaC platform.
\end{itemize}

Here is a small example \C program contained in file \verb|demo.c|:
\begin{shell}
int f(int i);
int g(int j);
void job(int *t, int A) {
  for(int i = 0; i < 50; i++) t[i] = f(i);
  for(int j = A; j < 100; j++) t[j] = g(j);
}
\end{shell}
and here is the content of the file \verb|demo.acsl|:
\begin{shell}
module demo :
 axiomatic A {
 type event = B | C(integer);
 predicate P(integer x);
 logic integer phi(event e);
 }

 function f:
contract:
  ensures P(phi(C(\result)));
  assigns \nothing;

function g:
contract: assigns \nothing;

function job:
at 1: assert 50 <= A <= 100;
at loop 1:
  loop invariant 0 <= i <= 50;
  loop invariant \forall integer k; 0 <= k < i ==> P(t[k]);
  loop assigns i,t[0..49];
at loop body 1:
  contract:
    ensures i == \old(i);
at loop stmt 1:
  contract:
    requires i==0;
    ensures  i==50;

at loop 2:
  loop invariant A <= j <= 100;
  loop assigns j,t[A..99];

at return:
  assert \forall integer k; 0 <= k < 50 ==> P(t[k]);

at call f:
  contract:
    ensures P(phi(C(t[i])));
\end{shell}
Here is the result of the \plugin{}:
\begin{shell}
# frama-c -acsl-import demo.acsl demo.c -print -no-unicode
[kernel] Parsing demo.c (with preprocessing)
[acsl-importer] Success for demo.acsl
[acsl-importer] Done: 1 file.
[kernel] Parsing demo.c (with preprocessing)
[acsl-importer] Success for demo.acsl
[acsl-importer] Done: 1 file.
/* Generated by Frama-C */
/*@
axiomatic A {
  type event = B | C(integer);
  predicate P(integer x) ;
  logic integer phi(event e) ;
  }
 */
/*@ ensures P(phi(C(\result)));
    assigns \nothing; */
int f(int i);

/*@ assigns \nothing; */
int g(int j);

void job(int *t, int A) {
  /*@ assert 50 <= A <= 100; */
  {
    int i = 0;
    /*@ requires i == 0;
        ensures i == 50; */
    /*@ loop invariant 0 <= i <= 50;
        loop invariant \forall integer k; 0 <= k < i ==> P(*(t + k));
        loop assigns i, *(t + (0 .. 49));
    */
    while (i < 50) {
      /*@ ensures i == \old(i); */
      {
        /*@ ensures P(phi(C(*(t + i)))); */
        *(t + i) = f(i);
      }
      i ++;
    }
  }
  {
    int j = A;
    /*@ loop invariant A <= j <= 100;
        loop assigns j, *(t + (A .. 99)); */
    while (j < 100) {
      {
        *(t + j) = g(j);
      }
      j ++;
    }
  }
  /*@ assert \forall integer k; 0 <= k < 50 ==> P(*(t + k)); */
  return;
}

int T[100];
void main(void) {
  job(T,50);
  return;
}
\end{shell}

\section*{Adding properties from the Graphical User Interface}
\markright{}
\addcontentsline{toc}{section}{Adding properties from the Graphical User Interface}
Contextual menus are added when selecting statements or function definitions.
It is possible to insert \verb|acsl_logical_axiomatic|, \verb|acsl_logical_type|,
           \verb|acsl_logical_predicate|, \verb|acsl_logical_function|,
           \verb|acsl_volatile_clause|, \verb|acsl_stmt_spec|,
           \verb|acsl_loop_annotation| and \verb|acsl_code_annotation|.
Selection of such a menu opens a popup window allowing to type directly the corresponding annotation (i.e. \verb|ensures P(t[0]);|).

\section*{Addons for \acsl}
\markright{}
\addcontentsline{toc}{section}{Addons for \acsl}
The option \verb|-acsl-import-addon-ensures-and-exits| adds a new clause \verb|ensures_and_exits| to \acsl grammar. This keyword is syntactic sugar for writing two clauses (one \verb|ensures| and one \verb|exits|) in one.

The option \verb|-acsl-import-addon-integer-cast| allows the introduction of casts from integer to C integral types. Warnings of the category \verb|annot:integer-cast| are emitted when introducing such casts.


\section*{Addons for code transformations}
\markright{}
\addcontentsline{toc}{section}{Addons for code transformation}
From \FramaC kernel, it is possible to identify two kinds of code transformations:
\begin{itemize}
\item transformations performed during the elaboration of the AST (Abstract Syntaxtic Tree) related to the parsed C files, and
\item transformations performed over the AST.
\end{itemize}

The import of ACSL specification files can be viewed as a second class transformation adding to the ACSL annotation to the current AST.
But, some imports need a code transformation of the first class for being performed (i.e. for identifying the statement body of the i$^{th}$ loop).
All syntaxic code transformations of \FramaC kernel (including the deletion of unused C symbols) are processed before the import of ACSL specification files.

The option \verb|-acsl-import-unroll-loop-conditions| transforms \verb|while| loops and and \verb|for| loops into \verb|do-while| loops when the initial loops have no loop annotation.

The \verb|while| loop
\begin{shell}
  while (c)
    //@ invariant I;
    S;
\end{shell}
is transformed into :
\begin{shell}
  if (c)
    do //@ invariant I;
      S;
    while (c);
\end{shell}

The \verb|for| loop
\begin{shell}
  for (init; c; incr)
    //@ invariant I;
    S;
\end{shell}
is transformed into :
\begin{shell}
  init ;
  if (c)
    do //@ invariant I;
    { S;
      incr;
    } while (c);
\end{shell}

For \verb|do-while| loops, \verb|invariant| clauses set on their loop body are equivalent to \verb|loop| \verb|invariant| clauses.
So, the importer is able to transform \verb|invariant| clauses set on every body of loops of every kind into \verb|loop invariant| clauses of \verb|do-while| loops.

This transformation is a first class transformation processed before the loop unrolling related to the option \verb|-kernel-ulevel| \verb|<n>|.
Since the code transformation processed by the option \verb|-kernel-ulevel| \verb|<n>| is also a first class transformation, the use of the options \verb|-kernel-ulevel| \verb|-1| {\tt <acsl-import-options>} \verb|-then| \verb|-acsl-import-run| \verb|-acsl-ulevel <n>| is required to perform loop unrolling after the import of ACSL specification files.

The option \verb|-acsl-import-ulevel <n>| unrolls loops as \verb|-kernel-ulevel <n>| does at the parsing stage, except that is now a second class transformation.
This means that a code annotation imported to a statement of the body of an unrolled loops can be duplicated using option \verb|-acsl-import-ulevel <n>| as if it was inside the C file and using the option \verb|-kernel-ulevel <n>| instead.
These \verb|UNROLL| loop pragmas could be in the source code, imported from ACSL specification files or generated by the {\tt\/-acsl-import-ulevel-spec} option.

The option {\tt\/-acsl-import-ulevel-spec <spec$_{1}$,...,spec$_{s}$>} adds \verb|UNROLL| loop pragmas that can be used by both  \verb|-acsl-import-ulevel| and \verb|-kernel-ulevel| transformations.
An unrolling specification \verb|<m@f:tag@n>| adds an annotation '{\tt\/loop unfold "tag", <n>;}' to the {\tt m$_{th}$} loop of the function \verb|<f>|.
An unrolling specification \verb|<c@f:tag@n>| adds an annotation '{\tt\/loop unfold "tag", <n>;}' to all loops of category \verb|<c>| of the function \verb|<f>|, where allowed loop categories are: \verb|while|, \verb|for| and \verb|do-while|.\\
A specification is considered as a set of elementary specifications: {\tt\/spec$_{1}$}, ..., {\tt\/spec$_{s}$}.\\
Categories, function names and loop occurrence numbers can be omitted.
The priority ordering used for choosing the (\verb|"tag"| name, unrolling value \verb|<n>|) pair is: \verb|<m@f:tag@n>| > \verb|<c@f:tag@n>| > \verb|<f:tag@n>| > \verb|<c:tag@n>| > \verb|<:tag@n>|.\\
The default value for optional tags is the empty string which leads to add unrolling loop pragmas without tags.\\
Nothing is done for loops having already a clause '{\tt\/loop unfold} ...'.\\


The command line
\begin{shell}
# frama-c -acsl-import-unroll-loop-conditions -ulevel-spec "fct:completely@2" file.c
\end{shell}
 will transforms the function
\begin{shell}
void fct (void) {
  S1;
  for (i=0; i<2; i++)
    Si;
  Sn ;
}
\end{shell}
such that the remaining loop have to be {\sl death} code:
\begin{shell}
void fct (void) {
  S1;
  i=0;
  if (i<2){ /*** loop condition unrolled ***/
    Si;     /*** first unrolling ***/
    i ++;
    if (! (i < 2)) goto unrolling_2_loop;
      Si;   /*** second unrolling ***/
      i ++;
      if (! (i < 2)) { goto unrolling_2_loop; }
      /*** the loop should be death code since completely unrolled ***/
      /*@ loop invariant \false;
          loop unfold "completely", 2; */
      while (1) {
        Si;
        i ++;
        if (! (i < 2)) break;
        }
unrolling_2_loop:
    }
  Sn ;
}
\end{shell}

\appendix
\include{user-changes}

\bibliographystyle{plain}
\bibliography{bib}
\addcontentsline{toc}{chapter}{\bibname}

\cleardoublepage
\end{document}
