\chapter*{Changes}
\addcontentsline{toc}{chapter}{Changes}

%\section*{Frama-C+dev}

\section*{Frama-C 32.0 (Germanium)}
\begin{itemize}
\item The plug-in is now part of the Frama-C open-source distribution
\item Options are now prefixed with \verb|-acsl-import|
\end{itemize}

\section*{From plugin ACSL importer 1.11.27 (Frama-C 27 Cobalt)}
\begin{itemize}
\item Minor changes for compatibility with Frama-C 27.
\end{itemize}

\section*{From plugin ACSL importer 1.11.26 (Frama-C 25 Iron)}
\begin{itemize}
\item The build system moves from \verb|Make| files to \verb|Dune| files.
\end{itemize}

\section*{From plugin ACSL importer 1.11.25 (Frama-C 25 Manganese)}
\begin{itemize}
\item Minor changes for compatibility with Frama-C 25.
\end{itemize}

\section*{From plugin ACSL importer 1.11.24 (Frama-C 24 Chromium)}
\begin{itemize}
\item Fixes import of the clause terminates.
\end{itemize}

\section*{From plugin ACSL importer 1.10.23 (Frama-C 23 Vanadium)}
\begin{itemize}
\item Fixes demo example (the global axiomatic to import was missing).
\end{itemize}

\section*{From plugin ACSL importer 1.9.19 (Frama-C 19 Potassium)}
\begin{itemize}
\item Fixes import of contracts related to labeled statements.
\item Fixes scope rule for searching variables under \verb|\at| construct.
\item Fixes individual status for \verb|ensures_and_exits| extended clause.
\end{itemize}

\section*{From plugin ACSL importer 1.8.18 (Frama-C 18 Argon)}
\begin{itemize}
\item Adding options \verb|-acsl-keep-unused-symbols|, \verb|-acsl-parse-only| and \verb|acsl-typing-only| options.
\end{itemize}

\section*{From plugin ACSL importer 1.7.17 (Frama-C 17 Chlorin}
\begin{itemize}
\item Extended grammar allowing the import of statement contracts and code annotations to loops.
\end{itemize}

\section*{From plugin ACSL importer 1.6.15 (Frama-C 15 Phosphorus)}
\begin{itemize}
\item Introduction of global macro.
\end{itemize}

\section*{From plugin ACSL importer 1.5.14 (Frama-C 14 Silicon)}
\begin{itemize}
\item Introduction of scopes for macro binding.
\end{itemize}

\section*{From plugin ACSL importer 1.4.13 (Frama-C 13 Aluminium)}
\begin{itemize}
\item Extended grammar allowing specification imports to C function call statements and assembly code.
\item Review of the grammar related to ``module'' directive.
\end{itemize}

\section*{From prom plugin ACSL importer 1.1.8 (Frama-C 8 Oxygen)}
\begin{itemize}
\item Section ``Addons for code transformation'' added.

\item Section ``Adding properties from the Graphical User Interface'' added.
\end{itemize}


\section*{From plugin ACSL importer 1.0.7 (Frama-C 7 Nitrogen) }
\begin{itemize}
\item Section ``Addons for \acsl'' added.

\item Initial document revision.
\end{itemize}
